---
title: "Decision Trees"
author: "INSOFE Lab Activity on Decision Trees"
date: "23 July 2017"
output:
  html_document:
    toc: yes
    toc_depth: 3
    toc_float: yes
  word_document:
    toc: yes
    toc_depth: '3'
---

# C5.0 Trees

**NOTE** Before starting this assignment please remember to clear your environment, you can do that by running the following code chunk

```{r}
rm(list = ls(all=T))

```

## Goal

* The goal of this activity is to predict wether a patient has liver disease or not based on various patient related attributes


## Agenda 

* Get the data

* Data Pre-processing

* Build a model

* Predictions

* Communication


## Reading & Understanding the Data

### Read the Data

Make sure the dataset is located in your current working directory, or else you can change your working directory using the "setwd()" function.

```{r}
getwd()

setwd("G:/DT assignment")

read.csv("ilpd_data.csv",header = T,sep = ",")->df
df
dim(df)
```

### Understand the data

* Use the str(), summary(), head() and tail() functions to get the dimensions and types of attributes in the dataset

* The dataset has 582 observations and 11 variables

- __The variable descriptions are given below:__

1 - age	 : Age of the patient

2 - gender : Gender of the patient

3 - TB : Total Bilirubin content

4 - DB : Direct Bilirubin content

5 - alk_phos : Alkaline Phosphotase content

6 - alamine : Alamine Aminotransferase content

7 - aspartate : Aspartate Aminotransferase content

8 - TP : Total Protiens content

9 - albumin : 	Albumin content

10 - A/G : Ratio of Albumin and Globulin 

11 - Disease : Whether the patient has liver disease or not 


```{r}
str(df)
summary(df)

```

```{r}
head(df)
tail(df)

```


## Data Pre-processing

### Verify Data Integrity

* Verify if the dataset has missing values

```{r}
sum(is.na(df))

colSums(is.na(df))
```

* Verify the data types assigned to the variables in the dataset

```{r}
str(df)
#lapply(df,table)

```

### Split the Data into train and test sets

* Use stratified sampling to split the data into train/test sets (70/30)

* Use the createDataPartition() function from the caret package to do stratified sampling

```{r}
library(caret)
set.seed(123)
createDataPartition(df$disease,p = 0.7,list = F)->dfsplit
df[dfsplit,]->trainlp
trainlp
dim(trainlp)

df[-dfsplit,]->testlp
dim(testlp)
colSums(is.na(trainlp))

```

### Impute the missing values

* Impute missing values using knnImputation() function in both the train and test datasets

```{r}

library(DMwR)


knnImputation(data = trainlp,k = 5)->trainlp
colSums(is.na(trainlp))

knnImputation(data = testlp,k=5)->testlp
colSums(is.na(testlp))

```

## Build a  Decision Tree

### Model the tree

* Use Quinlan's C5.0 decision tree algorithm implementation from the C50 package to build your decision tree

```{r}
library(C50)

C5.0(disease~.,data = trainlp)->dftree
dftree


```

* Build a rules based tree

```{r}
C5.0(disease~.,data = trainlp,rules=T)->dfrules
dfrules


```

### Variable Importance in trees

* Find the importance of each variable in the dataset

```{r}

#C5imp(dftree,metric = "usage")
#C5imp(dftree,metric = "usage")
```

### Rules from trees

* Understand the summary of the returned c5.0 rules based on the decision tree model


```{r}
summary(dfrules)


```


### Plotting the tree

* Call the plot function on the tree object to visualize the tree

```{r, fig.width= 35, fig.height=15}

plot(dftree)
```


## Evaluating the model

### Predictions on the test data

* Evaluate the decision tree using the standard error metrics on test data

```{r}

predict(dftree,testlp)->dfpredt

dfpredt

```

* Report error metrics for classification on test data

```{r}

library(caret)
#library(rattle)
confusionMatrix(dfpredt,testlp$disease)


```

# CART Trees

**NOTE** Before starting this assignment please remember to clear your environment, you can do that by running the following code chunk

```{r}

rm(list = ls(all=T))

```

* The classification and regression trees use gini index in place of the gain ratio (based on information gain) used by the ID3 based algorithms, such as c4.5 and c5.0

## Goal

* The goal of this activity is to predict the heating load of a residential building, if the building parameters are given

* Hence, in the future architects would be able to build more energy efficient buildings as they can optimize the building parameters to reduce the heating load

## Agenda 

* Get the data

* Data Pre-processing

* Build a model

* Predictions

* Communication


## Reading & Understanding the Data

### Read the Data

* Make sure the dataset is located in your current working directory, or else you can change your working directory using the "setwd()" function.

```{r}
getwd()
setwd("G:/DT assignment")

read.csv("building_energy.csv",header = T,sep = ",")->be_data
be_data
dim(be_data)

```

### Understand the data

* Use the str(), summary(), head() and tail() functions to get the dimensions and types of attributes in the dataset

* The dataset has 768 observations and 9 variables

```{r}
str(be_data)
summary(be_data)
head(be_data)
tail(be_data)
```

* The variable names are self explanatory, for further information visit http://www.sciencedirect.com/science/article/pii/S037877881200151X

## Data Pre-processing

### Verify Data Integrity

* Verify if the dataset has missing values

```{r}
sum(is.na(be_data))


```

* Verify the data types assigned to the variables in the dataset

```{r}

# Enter answer here
str(be_data)


```

### Split the Data

* Split the data into train/test sets (70/30)

```{r}
set.seed(123)
sample(1:nrow(be_data),0.7*nrow(be_data))->be_split
be_split

be_data[be_split,]->be_train
be_train
dim(be_train)
be_data[-be_split,]->betest
dim(betest)

library(vegan)
  
```

## Build a Regression Tree

### Model the tree

* Use the rpart package to build a cart tree to predict the heating load

```{r}

library(rpart)

rpart(heating_load~.,data = be_train)->be_rpart
be_rpart
printcp(be_rpart)
```

### Tree Explicability

* Print the variable importance

```{r}

be_rpart$variable.importance

```

* Plot the regression tree

```{r, fig.width=8, fig.height=5}

library(rpart.plot)

rpart.plot(be_rpart)

```

## Evaluation on Test Data

* Report error metrics on the test data

```{r}

predict(be_rpart,betest)->pred
pred
```

```{r}

library(DMwR)
regr.eval(betest$heating_load,pred)

```




















